\ZpL: a p-adic precision package

### 2018-02-03 ZpL/ZpL.tex
More comments in Section 2

\newcommand{\sage}{\textsc{SageMath}\xspace}
\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}

@@

We begin our tour of the features of the \ZpL package
with some basic arithmetic computations.
The first step is the definition of the $p$-adic ring or field.

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
 & {\color{ring}\verb?Z2?}\verb? = Zp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
 & {\color{ring}\verb?Q2?}\verb? = Qp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
\end{tabular}}

@@

\noindent
It means that we will work in various implementation of the 
ring of $2$-adic integers with $20$ digits of precision (which
is the default).
The rings {\color{ring}\verb?Z3?} and {\color{ring}\verb?Q3?} are defined similarly.

We show the results for \ZpCR%, \ZpFP 
and \ZpLC.
Results for \ZpLF are only shown when they differ from \ZpLC.

@@

We first take some random element $x$.
\todo{Should we say a word about the function \texttt{random\_element}?}

@@

Multiplication by $3$ is a shift on the digits, whether it 
is done by addition or multiplication.

@@

\noindent
\todo{Add a short explanation?}

The same phenomenon occurs for multiplication.
%when comparing the computation of $x^3$ and $x \times x^2$.

@@

\noindent
No loss in precision appears in \ZpLC, contrary to \ZpCR.
%We remark that \ZpFP also provides all the correct digits.

@@

In linear algebra, determinant and characteristic polynomial are 
notoriously hard to compute: see \cite{caruso-roe-vaccon:15, 
caruso-roe-vaccon:17}. We illustrate this with the following example 
(using a division-free algorithm for the computation of the 
characteristic polynomial) :

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
 & \verb?M.?{\color{method}\verb?charpoly?}\verb?()? \\
\cZpCR
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
\cZpLC
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...11100111.01*x +? \\
 & \verb?...010000010? \\
\end{tabular}}

@@

@@

Our package can be applied to computation over $p$-adic polynomials. A 
natural example is that of the computation of GCD, whose stability has 
been studied in \cite{caruso:2017}.

@@

\noindent
With high probability, $P$ and $Q$ are coprime, implying that the gcd of 
$DP$ is $DQ$ is $D$.  However, we observe that we do not always get this 
expected result. More precisely, \ZpCR often finds a gcd of higher 
degree. Indeed the Euclidean algorithm stops prematurely because the 
software believes that some early remainder vanishes due to the lack of 
precision.
%The situation with ZpFP is the exact opposite:
%since the precision is too high (always the maximal cap), 
%the software sometimes does not notice that the remainder vanishes.
%Hence, it happens that the Euclidean algorithm 
%obtains a last non-zero remainder of degree less than that of
%$D.$

>>>>

\newcommand{\sage}{\textsc{SageMath}\xspace}
\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
\newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
\newcommand{\QpXX}{\text{\color{output} \rm \tt QpXX}\xspace}

@@

The first step is to define the parents, that are the rings of $p$-adic
numbers we will work with.

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
 & {\color{ring}\verb?Z2?}\verb? = ?{\color{output}\verb?ZpXX?}\verb?(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
 & {\color{ring}\verb?Q2?}\verb? = ?{\color{output}\verb?QpXX?}\verb?(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
\end{tabular}}

@@

\noindent
\ZpXX is a generic notation for \ZpCR, \ZpLC and \ZpLF.
The first one \ZpCR is the usual constructor for $p$-adic parents in 
\sage. It tracks precision using interval arithmetics. On the contrary
\ZpLC and \ZpLF are provided by our package. In the sequel, we will
compare the results output with the three 
Results for \ZpLF are only shown when they differ from \ZpLC.

@@

We begin our tour of the features of the \ZpL package with some basic 
arithmetic computations.
We first pick some random element $x$.
The function {\color{function}\verb?random_element?} is designed so
that it guarantees that the picked random element is the same for
each constructor \ZpCR, \ZpLC and \ZpLF.

@@

\noindent
Multiplication by $p$ (here $3$) is a shift on the digits and thus
leads to a gain of one digit in absolute precision.
On the example below, we observe that when this multiplication is
split into several steps, \ZpCR does not see the gain of precision
while \ZpLC and \ZpLF do.

@@

\noindent
The same phenomenon occurs for multiplication.
%when comparing the computation of $x^3$ and $x \times x^2$.

@@

\noindent
Again \ZpCR does not output the optimal precision when the computation
is split into several steps whereas \ZpLC and \ZpLF do.
Actually, these toy examples illustrate quite common situations which 
often occur during the execution of many algorithms. For this reason,
interval arithmetics often overestimates the losses of precision. 
Roughly speaking, the aim of our package is to ``fix this misfeature''.
In the next subsections, we present a bunch of examples showing the
benefit of \ZpL in various contexts.

@@

In linear algebra, characteristic polynomials are notoriously hard to 
compute: see \cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We 
illustrate this with the following example (using a division-free 
algorithm for the computation of the characteristic polynomial) :

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
 & \verb?M.?{\color{method}\verb?determinant?}\verb?()? \\
\cZpCR
 & \verb?...010000010? \\
\cZpLC
 & \verb?...010000010? \\
\cIn
 & \verb?M.?{\color{method}\verb?charpoly?}\verb?()? \\
\cZpCR
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
\cZpLC
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...11100111.01*x +? \\
 & \verb?...010000010? \\
\end{tabular}}

@@

\noindent
We observe that \ZpLC can guarantee $4$ more digits on the $x$ 
coefficient. Moreover, it recovers the correct precision on the
constant coefficient (which is the determinant) whereas \ZpCR is 
confused and cannot even certify that it does not vanish.

@@

Our package can be applied to computation with $p$-adic polynomials. A 
natural example is that of the computation of GCD, whose stability has 
been studied in \cite{caruso:2017}.

@@

\noindent
With high probability, $P$ and $Q$ are coprime, implying that the gcd of 
$DP$ is $DQ$ is $D$.  However, we observe that \ZpCR output a quite 
different result. The point is that, in the \ZpCR case, Euclidean 
algorithm stops prematurely because the test \verb?B != 0? fails 
too early due to the lack of precision.
%The situation with ZpFP is the exact opposite:
%since the precision is too high (always the maximal cap), 
%the software sometimes does not notice that the remainder vanishes.
%Hence, it happens that the Euclidean algorithm 
%obtains a last non-zero remainder of degree less than that of
%$D.$
